Topology and experimental distinguishability

### 2017-07-27 Articles/TopologyAndExperimentalDistinguishability/article.tex
Incorporating Christine's comments in the main body

In science, a statement can be accepted as true only if there exists a way to independently verify it. To capture this notion we introduce the following definitions.

@@

\begin{defn}
Any experimental observation whose statement is a contradiction is also called \textbf{contradiction} and is noted by $\bot$.
\end{defn}

@@

\begin{rem}
	We do allow infinite time for the verification of a domain with the understanding that some domains will only be approximately verified in finite time. As we have, so to speak, only one infinity to spend, we spend it here to maximize its usefulness.
\end{rem}

@@

\begin{defn}
	A \textbf{sub-basis} of an experimental domain is any subset that can generate all others via finite conjunction and countable disjunction. A \textbf{basis} of an experimental domain is any subset that can generate all others by countable disjunctions. 
\end{defn}

\begin{rem}
	As for topologies, given a sub-basis one can generate a basis by taking all finite conjunctions of the observations. Any infinite sub-basis will generate a basis of the same cardinality.
\end{rem}

\begin{prop}
Let $\mathcal{D}$ be an experimental domain. Then there exists a countable basis (equivalently, sub-basis) $\mathcal{B}$ of $\mathcal{S}$.
\end{prop}

@@

If there does not exist a countable basis, then by definition there does not exist a sequence of experimental observations in $\mathcal{D}$ from which one can deduce all other observations in $\mathcal{D}$. Hence it is impossible to test all members of $\mathcal{S}$.
\end{proof}

@@

We now turn our attention to a more specific case. We want to characterize an experimental domain whose function is to identify the value of an attribute among all possible values.

\begin{defn}
	An \textbf{attribute} is the triple $(X, x, \mathcal{D}_x)$ where:
	\begin{itemize}
		\item $X$ is the set of all possible values, which we call \textbf{possibilities}, and satisfies $|X|>1$
		\item $x$ is the value of the attribute, therefore $x \in X$
		\item $\mathcal{D}_x$ is an experimental domain containing all possible experimental observations of the form $\mathsf{o} = (x\in U, \mathsf{e}_\in(U))$ where $U \subseteq X$ is a set of possibilities and $\mathsf{e}_\in(U)$ is an experimental test that succeeds if and only if $x \in U$
	\end{itemize}
	Any subset $U\subseteq X$ for which such an observation exists is said to be a \textbf{verifiable set}.
\end{defn}

\begin{lem}
\label{setbehavior}
	Let $U_1, U_2, ... , U_n, ...$ be a countable infinite sequence of verifiable sets. The finite intersection $\bigcap\limits_{i=1}^{n} U_i$ and the countable union $\bigcup\limits_{i=1}^{\infty} U_i$ are verifiable sets.
\end{lem}

@@

\begin{defn}
The possibilities of an attribute $(X, x, \mathcal{D}_x)$, and by extension the attribute itself, are said \textbf{experimentally distinguishable} if $\mathcal{D}_x$ is such that given any two possibilities $x_1, x_2 \in X$ we can find two incompatible experimental observations $(x\in U_1, \mathsf{e}_\in(U_1)), (x\in U_2, \mathsf{e}_\in(U_2))\in\mathcal{D}$ such that $x_i\in U_i$ for $i=1,2$. 
\end{defn}

@@

\begin{thm}
The set of experimentally distinguishable possibilities for an attribute $(X, x, \mathcal{D}_x)$ is a Hausdorff, second-countable topological space $(X,\mathsf{T})$ where the open sets are given by the verifiable sets. 
\end{thm}
\begin{proof}
First, from the definition one can see that for all $x\in X$, there exists a verifiable set $U$ with $x\in U$, so their union $X$ is also a verifiable set. Further, there is a null observation by taking the conjunction of two identifications corresponding to disjoint sets, so the empty set is a verifiable set. Now, Proposition \ref{setbehavior} shows that the collection $\mathsf{T}$ is closed under finite intersection and countable union. Because $\mathsf{T}$ is determined by an experimental domain of observations, there is a countable basis of operations which translates to a countable basis of open (verifiable) sets, so it is second-countable. To show it is closed under arbitrary union, notice that an arbitrary union may be rewritten as a union of basis elements, which is then a countable union, and so it remains in $\mathsf{T}$. That the topology is Hausdorff is immediate from the last part of the definition. 
\end{proof}

\begin{rem}
As any Hausdorff, second-countable topological space has at most cardinality of the continuum, that is also the greatest cardinality that a set of experimentally distinguishable objects can have. We can conclude that set of mathematical objects, such as all functions from $\mathbb{R}$ to $\mathbb{R}$, that do not satisfy this requirement are not good candidates to represent scientific concepts.
\end{rem}

@@

We now want to characterize relationships between two attributes. Such relationships can be defined either on the values (i.e. the possibilities) or on the observations (i.e. the experimental domain). We need to show that both definitions lead to the same mathematical object.

\begin{defn}[Experimental relationship on possibilities]
	Let $(X, x, \mathcal{D}_x)$ and $(Y, y, \mathcal{D}_y)$ be two attributes. An \textbf{experimental relationship} is a map $f : X \rightarrow Y$ that can be used within an experimental test.
\end{defn}

@@

Next, we consider the opposite problem. Suppose we have two experimentally distinguishable spaces $(X,\mathsf{T}_X)$ and $(Y,\mathsf{T}_Y)$, and that we have a well-behaved (with respect to conjunction and disjunction) map on the experimental domains. Then this translates to a map of topologies, taking the verifiable (open) sets of $Y$ to those of $X$, which respects (arbitrary) union and (finite) intersection. In this case, we will show that one can find a continuous map which induces this map of topologies.

\begin{defn}[Experimental relationship on observations]
	Let $(X, x, \mathcal{D}_x)$ and $(Y, y, \mathcal{D}_y)$ be two attributes. An \textbf{experimental relationship} is a map $g : \mathcal{D}_y \rightarrow \mathcal{D}_x$ such that if $\mathsf{o}_y \in \mathcal{D}_y$ is verified then $\mathsf{o}_x \in \mathcal{D}_x$ must also be verified. To be consistent, such relationship must have these properties:
	\begin{enumerate}
	\item it is compatible with conjunction and disjunction: for any $\mathsf{o}_1, \mathsf{o}_2 \in \mathcal{D}_y$, we have $g(\mathsf{o}_1 \wedge \mathsf{o}_2)=g(\mathsf{o}_1)\wedge g(\mathsf{o}_2)$ and $g(\mathsf{o}_1 \vee \mathsf{o}_2)=g(\mathsf{o}_1)\vee g(\mathsf{o}_2)$
	\item contradiction leads to contradiction: $g(\bot) = \bot$
	\item no knowledge leads to no knowledge: if $Y$ is the verifiable set associated with $\mathsf{o}_y$ then $X$ is the verifiable set associated with  $g(\mathsf{o}_y)$
	\end{enumerate}
\end{defn}

\begin{prop}
	Each experimental relationship $g$ defined above is associated with a unique continuous function $f: X \rightarrow Y$ such that $g(\llparenthesis y\in U_Y,\mathsf{e}_{\in}(U_Y)\rrparenthesis) = \llparenthesis x\in f^{-1}(U_Y),\mathsf{e}_{\in}(f^{-1}(U_Y))\rrparenthesis$.
\end{prop}

@@

\begin{thm}
	An experimental relationship between two sets $X$ and $Y$ of experimentally distinguishable possibilities is a continuous function $f : X \rightarrow Y$ with respect to the between the respective topological spaces $(X,\mathsf{T}_X)$ and $(Y,\mathsf{T}_Y)$ given by the verifiable sets. 
\end{thm}

\begin{rem}
	This result gives a formal justification as to why continuous functions are prevalent in science in general and in physics in particular. As topologies capture experimental distinguishability, continuous function preserve it.
\end{rem}

@@

\begin{defn} Let $X$ and $Y$ be two topological spaces. Let $C(X,Y)$ denote the set of all continuous functions from $X$ to $Y$. Let $\mathcal{B}_X$ and $\mathcal{B}_Y$ be two bases for $X$ and $Y$ respectively. The basis-to-basis topology $\mathsf{T}(C(X,Y), \mathcal{B}_X, \mathcal{B}_Y)$ on $C(X,Y)$ with respect to the basis $\mathcal{B}_X$ and $\mathcal{B}_Y$ is the topology generated by all sets of the form 
	$$
	V(U_X, U_Y) = \{f\in C(X,Y) : f(U_X)\subset U_Y\}
	$$
where $U_X \in \mathcal{B}_X$ and $U_Y \in \mathcal{B}_Y$.
\end{defn}

Now we need to show that if the two spaces $X$ and $Y$ are Hausdorff and second-countable, the basis-to-basis topology is Hausdorff and second-countable for a suitable choice of basis.

@@

\begin{rem}
	Note that the basis-to-basis topology is not in general equal to the open-open topology. The former may depend on the basis chosen while the second is uniquely defined by the topology.
\end{rem}

As experimental relationships are themselves distinguishable, we can recursively form experimental relationships between experimental relationship leading to functions of arbitrary order while remaining within the definition provided. The framework is therefore complete.

>>>>

In science, a statement can be accepted as true only if there exists a way to experimentally verify it. To capture this notion we introduce the following definitions.

@@

\begin{defn}
Any experimental observation whose statement is a contradiction is also called a \textbf{contradiction} and is noted by $\bot$.
\end{defn}

@@

\begin{rem}
	We do allow infinite time for the verification of a domain with the understanding that some domains will only be partially verified in finite time. As we have, so to speak, only one infinity to spend, we spend it here to maximize its utility.
\end{rem}

@@

\begin{defn}
	A \textbf{sub-basis} of an experimental domain is any subset that can generate all others via finite conjunction and countable disjunction. A \textbf{basis} of an experimental domain is any subset that can generate all others by countable disjunction.
\end{defn}

\begin{rem}
	As for topologies, given a sub-basis one can generate a basis by taking all finite conjunctions. Any infinite sub-basis will generate a basis of the same cardinality.
\end{rem}

\begin{prop}
Let $\mathcal{D}$ be an experimental domain. Then there exists a countable basis (equivalently, sub-basis) $\mathcal{B}$ of $\mathcal{D}$.
\end{prop}

@@

If there does not exist a countable basis, then by definition there does not exist a sequence of experimental observations in $\mathcal{D}$ from which one can deduce all other observations in $\mathcal{D}$. Hence it is impossible to test all members of $\mathcal{D}$.
\end{proof}

@@

We now turn our attention to a more specific case. We want to characterize an experimental domain whose purpose is to identify an element among a set of possibilities.

\begin{defn}
	An \textbf{experimental identification domain} is the triplet $(\mathcal{D}_x, X, x)$ where:
	\begin{itemize}
		\item $X$ is the set of \textbf{possibilities}, and satisfies $|X|>1$
		\item $x$ is the element to identify among the possibilities, therefore $x \in X$
		\item $\mathcal{D}_x$ is an experimental domain containing all possible experimental observations of the form $\mathsf{o} = (x\in U, \mathsf{e}_\in(U))$ where $U \subseteq X$ is a set of possibilities and $\mathsf{e}_\in(U)$ is an experimental test that succeeds if and only if $x \in U$
	\end{itemize}
	Any subset $U\subseteq X$ for which such an observation exists is said to be a \textbf{verifiable set}.
\end{defn}

\begin{lem}
\label{setbehavior}
	Let $U_1, U_2, ... , U_n, ...$ be a countably infinite sequence of verifiable sets. The finite intersection $\bigcap\limits_{i=1}^{n} U_i$ and the countable union $\bigcup\limits_{i=1}^{\infty} U_i$ are verifiable sets.
\end{lem}

@@

\begin{defn}
A set of possibilities $X$ is \textbf{experimentally distinguishable} if there exists an experimental identification domain $(\mathcal{D}_x, X, x)$ such that for any two possibilities $x_1, x_2 \in X$ we can find two incompatible experimental observations $(x\in U_1, \mathsf{e}_\in(U_1)), (x\in U_2, \mathsf{e}_\in(U_2))\in\mathcal{D}$ such that $x_i\in U_i$ for $i=1,2$. 
\end{defn}

@@

\begin{thm}
A set of experimentally distinguishable possibilities $X$ has a natural Hausdorff, second-countable topology $(X,\mathsf{T})$ with the open sets given by the verifiable sets of the associated experimental identification domain $(\mathcal{D}_x, X, x)$.
\end{thm}
\begin{proof}
First, from the definition one can see that for all $x\in X$, there exists a verifiable set $U$ with $x\in U$, so their union $X$ is also a verifiable set. Further, there is a null observation by taking the conjunction of two identifications corresponding to disjoint sets, so the empty set is a verifiable set. Now, Proposition \ref{setbehavior} shows that the collection $\mathsf{T}$ is closed under finite intersection and countable union. Because $\mathsf{T}$ is determined by an experimental domain of observations, there is a countable basis of operations which translates to a countable basis of open (verifiable) sets, so it is second-countable. To show it is closed under arbitrary union, notice that an arbitrary union may be rewritten as a union of basis elements, which is then a countable union, and so it remains in $\mathsf{T}$. That the topology is Hausdorff is immediate from the last part of the definition. 
\end{proof}

\begin{rem}
As any Hausdorff, second-countable topological space has at most cardinality of the continuum, that is also the greatest cardinality that a set of experimentally distinguishable objects can have. We can conclude that sets of mathematical objects, such as all functions from $\mathbb{R}$ to $\mathbb{R}$, that do not satisfy this requirement are not good candidates to represent scientific concepts.
\end{rem}

@@

We now want to characterize relationships between two experimentally distinguishable elements. Such relationships can be defined either on the values (i.e.~the possibilities) or on the observations (i.e.~the experimental domain). We need to show that both definitions lead to the same mathematical object.

\begin{defn}[Experimental relationship between possibilities]
	Let $(\mathcal{D}_x, X, x)$ and $(\mathcal{D}_y, Y, y)$ be two experimental identification domains. An \textbf{experimental relationship} is a map $f : X \rightarrow Y$ that can be used within an experimental test.
\end{defn}

@@

\begin{defn}[Experimental relationship between observations]
	Let $(\mathcal{D}_x, X, x)$ and $(\mathcal{D}_y, Y, y)$ be two experimental identification domains. An \textbf{experimental relationship} is a map $g : \mathcal{D}_y \rightarrow \mathcal{D}_x$ such that if $\mathsf{o}_y \in \mathcal{D}_y$ is verified then $\mathsf{o}_x \in \mathcal{D}_x$ must also be verified. To be consistent, such relationship must have these properties:
	\begin{enumerate}
	\item it is compatible with conjunction and disjunction: for any $\mathsf{o}_1, \mathsf{o}_2 \in \mathcal{D}_y$, we have $g(\mathsf{o}_1 \wedge \mathsf{o}_2)=g(\mathsf{o}_1)\wedge g(\mathsf{o}_2)$ and $g(\mathsf{o}_1 \vee \mathsf{o}_2)=g(\mathsf{o}_1)\vee g(\mathsf{o}_2)$
	\item contradiction leads to contradiction: $g(\bot) = \bot$
	\item no knowledge leads to no knowledge: if $Y$ is the verifiable set associated with $\mathsf{o}_y$ then $X$ is the verifiable set associated with  $g(\mathsf{o}_y)$
	\end{enumerate}
\end{defn}

\begin{prop}
	For each experimental relationship $g$ defined above there exists a unique continuous function $f: X \rightarrow Y$ such that $g(\llparenthesis y\in U_Y,\mathsf{e}_{\in}(U_Y)\rrparenthesis) = \llparenthesis x\in f^{-1}(U_Y),\mathsf{e}_{\in}(f^{-1}(U_Y))\rrparenthesis$.
\end{prop}

@@

\begin{thm}
	An experimental relationship between two sets $X$ and $Y$ of experimentally distinguishable possibilities is a continuous function $f : X \rightarrow Y$ between the respective natural topologies $(X,\mathsf{T}_X)$ and $(Y,\mathsf{T}_Y)$. 
\end{thm}

\begin{rem}
	This result gives a formal justification as to why continuous functions are prevalent in science in general and in physics in particular. As topologies capture experimental distinguishability, continuous functions preserve it.
\end{rem}

@@

\begin{defn} Let $X$ and $Y$ be two topological spaces. Let $C(X,Y)$ denote the set of all continuous functions from $X$ to $Y$. Let $\mathcal{B}_X$ and $\mathcal{B}_Y$ be two bases for $X$ and $Y$ respectively. The \textbf{basis-to-basis topology} $\mathsf{T}(C(X,Y), \mathcal{B}_X, \mathcal{B}_Y)$ on $C(X,Y)$ with respect to the basis $\mathcal{B}_X$ and $\mathcal{B}_Y$ is the topology generated by all sets of the form 
	$$
	V(U_X, U_Y) = \{f\in C(X,Y) : f(U_X)\subset U_Y\}
	$$
where $U_X \in \mathcal{B}_X$ and $U_Y \in \mathcal{B}_Y$.
\end{defn}

@@

\begin{rem}
	Note that the basis-to-basis topology is not in general equal to the open-open topology. The former may depend on the choice of bases $\mathcal{B}_X$ and $\mathcal{B}_Y$ while the second is uniquely defined by the topologies of $X$ and $Y$.
\end{rem}

As experimental relationships are themselves distinguishable, we can recursively form experimental relationships between experimental relationships leading to functions of arbitrary order while remaining within the definitions provided. The framework is therefore complete.
